Prenatal conditions do not affect brain physiology and learning in a lizard

### 2025-04-11 ms.qmd
updates

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen^3^, Daniel W.A. Noble^1^

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ Flinder's University, College of Science and Engineering, Bedford Park, SA 5042, Australia  
^3^ University of Wollongong, Wollongong, NSW 2500, Australia  
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

ORCID:

@@

## Abstract
Early environmental factors like heat or stress hormones can impair learning through brain metabolic function, crucial for neural development and synaptic plasticity. However, whether early environments always result in cognitive impairment through changes in neural physiology is not well-established outside of a few model systems. Here, we investigated the effects of prenatal temperature and corticosterone (CORT) on brain mitochondrial activity and spatial learning in the delicate skink (_Lampropholis delicata_). We treated eggs with either CORT or a control vehicle and incubated at cold (23 ± 3°C) or hot (28 ± 3°C) temperatures. Juveniles were tested in a spatial learning task over 40 days after which mitochondrial function in the medial cortex was assessed. Despite among-individual variation in learning ability, mitochondrial physiology and spatial learning in _L. delicata_ remained robust to prenatal temperature and CORT exposure. No significant relationship was found between mitochondrial function and cognitive performance, contrary to predictions. Additionally, increased metabolic capacity correlated with higher ROS production but did not affect oxidative damage, possibly due to protective mechanisms. These findings highlight the physiological and cognitive resilience of _L. delicata_ to early-life challenges. Future research should explore whether this robustness extends to other brain regions, cognitive domains, and life stages.

@@

Learning - the acquisition and consolidation of new information [@dukas_evolutionary_2004] - enables animals to create new associations between events, which can be essential for coping with environmental change [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form these associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Learning faster may imply exploiting novel resources better or responding to new threats more efficiently [@ward2016ecological]. Consequently, individual differences in learning abilities can have profound effects on fitness, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Understanding the mechanisms that drive these differences is therefore crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early-life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Adverse developmental environments are particularly influential, as the brain is highly sensitive to environmental inputs during early life [@zhu_prenatal_2004]. In fact, early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity is considered a key factor [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved in learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress - a byproduct of mitochondrial metabolism [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on mammals have shown the pervasive effects of mitochondrial physiology on cognitive abilities [@zhu_prenatal_2004; @hara_presynaptic_2014; @cao2014ampk]. However, the extent to which these effects are generalizable to other taxa remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. Maternal stress can also influence how mitochondria operate in offspring [@zhu_prenatal_2004]. Stressful situations experienced by mothers can elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can be transmitted to the developing embryos [@uller_sex-specific_2009] altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014]. For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) with impacts on spatial learning abilities [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Temperature is also a significant source of maternal stress in ectotherms. Thermal environments outside of the optimal range can elevate GCs in mothers, which can be passed to the offspring [see @Crino_2023]. Additionally, temperature can also directly influence offspring development, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment plays a crucial role in shaping mitochondrial function, affecting energy metabolism and oxidative stress [@stier2022experimental; @crino2024eggs]. Thus, the combined effects of prenatal GCs and temperature may profoundly influence mitochondrial function, with important consequences for brain development and cognition. However, the extent to which prenatal GCs and temperature interact to shape cognitive abilities via mitochondrial physiology remains largely unknown outside of a few model species [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities through impacts on brain mitochondrial activity. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs], especially if CORT can alter cellular components increasing ROS production without enhancing ATP synthesis or making them more vulnerable to oxidative damage. Similarly, high temperatures would decrease mitochondrial efficiency [@zavorka_climate_2021; @crino2024eggs], but with the potential to reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, that could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, shedding light on the role of early-life conditions in cognitive development and associated mechanismsm.

@@

#### Experimental animals
Lizards came from a breeding colony established in the laboratory in 2019 from wild populations in Sydney, Australia. The colony consisted of 270 adults living in groups of two males and four females. Eggs were collected from these groups between November 2023 to January 2024. After collecting the eggs, we treated them with CORT or a vehicle control and incubated them under two different temperature regimes (see below). Clutch and egg identity were assigned immediately after egg collection, and the eggs were incubated in individual cups until hatching. Hatchlings were also kept in individual enclosures until the end of the experiment. For details on husbandry and breeding conditions see _Methods 1: Animal husbandry in Supplementary Material_.

@@

#### Spatial learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact. In each trial, lizards were placed by hand in the center of maze and left to acclimatise for two minutes. During acclimatisation, the central area was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush was used to encourage movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

We assessed lizards' spatial learning using external cues for navigation while avoiding intra-maze cues. Subtle intra-maze cues were avoided by replacing the maze every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. Additionally, the maze was cleaned with 70% ethanol between trials to avoid the influence of chemical cues. To control for potential side biases, the correct arm was randomly assigned to one of the six arms for each trial. We employed four different maze orientations counterbalancing the number of lizards assigned to each orientation across treatments.

The task was repeated once daily for 40 consecutive days, and the number of errors made by each lizard was recorded. An error was defined as the lizard inserting its head into one of the incorrect arms, choosing the same arm two or more times consecutively was counted as independent errors.

#### Brain mitochondrial function
Immediately following the completion of the tests, we quantified mitochondrial physiology in the brain of lizards. We euthanized lizards via peritoneal injection 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation we evaluated lizard righting responses and the pinching reflexes in one of the front limbs.

@@

The aliquots reserved for oxidative damage assays were stained with 20 µL of 10 µg/mL Hoechst 33342 Nuclear Viability Dye and 20 µL of 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The samples were then fixed in 1% Neutral-Buffered Formalin, washed, and preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized in 200 µL of 1X PBS containing 5mM (6.14 mg/mL) digitonin. Following permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody - a marker of DNA damage - and allowed them to incubate overnight (~12 hours). The following day we counterstained the cells with 20µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyzed the samples using the flow cytometer. Oxidative damage assays were performed within 6 months of the initial analysis of fresh samples.

Flow cytometry was performed on a Becton Dickson LSRFortessa X-20 flow cytometer using the default wavelength filters. Detectors and voltage settings for each assay were determined during pilot trials and remained consistent throughout the experiment. Data was processed using FlowJo (v. 10.1) software. We obtained the mean fluorescent intensity for mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays see _Methods 2: flow cytometry in Supplementary Material_. Sample sizes for DNA damage and lipid peroxidation were smaller than for the other variables because of a plater reader malfunction, however, data was missing randomly and we dealt with missing data in our analyses (see below).

#### Statistical analyses
We performed all analyses using the _brm_ package [@burkner2017brms] which fits Bayesian multilevel models with Stan [@stan] with R version 4.4.0 [@R]. We ran a series of univariate and multivariate models to test the effects of early envinonment on each variable separately plus quantify the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warmup interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each of the variables recorded: mitochondrial density, metabolic capacity, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. We included in all the models the hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot) and their interaction. Sex and age were included in preliminary models and excluded from the final models when they were not significant. For all univariate models, clutch identity was included as a random factor. Mitochondrial related variables where log-transformed when necessary and all were mean-centered and standardized by dividing them by two times the standard deviation [@gelman2008scaling]. These variables were considered to follow a normal distribution.

Learning was modeled as a function of trial, cort, temperature, and their three-way interaction. We included in this model the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates less than 0 provide evidence of learning. _pMCMC_ was used to test the hypothesis that posterior distributions of slopes, and slope contrasts, were different from zero. We considered an effect statistically significant when _pMCMC_ <0.05.

To explore direct and indirect links between mitochondrial function and learning, we fitted a multivariate structural equation model (SEM) using _brms_. The model was structured based on theoretical expectations shown in @fig-sem. Any missing data was imputed during model fitting using data augmentation [@noble2021planned], but this was largely restricted to DNA damage and lipid peroxidation. To obtain a measure of learning performance we extracted the posterior distribution of individual learning slopes (i.e., changes in error over time) for each lizard. This was achieved by including a random trial slope for each lizard. Learning slopes, as well as the rest of variables, were standardized as before (see above). We assumed a gaussian error distribution for all the variables in this model. Factors found as non-significant in the univariate models were excluded. Clutch identity remained a random factor in our SEM model. Direct, indirect, and total effects were derived from posterior estimates [@kline2005principles] (see table S4).

@@

We started with 80 lizards - 20 per treatment - coming from `r total_clutches` clutches. However, due to natural mortality, only 79 lizards were included in the learning analyses, and 78 in the mitochondrial analyses. We provide final sample sizes in all our figures.

@@

```{r, fig-learning}
#| label: fig-learning
#| fig.cap: "Results for learning analyses. (a) the predicted number of errors over trials. The lines represent the mean predicted number of errors for each trial, and the shaded areas indicate the standard error of the mean; both were obtained using the slope and intercept estimates from the posterior distributions. (b) shows the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. Points and bars represent the mean and standard error of the estimated slopes, respectively. Dashed lines indicate value 0. The different colors in both panels indicate the different treatments. Asterisks in (b) indicate significant differences from 0."
source(here("R", "func.R"))
#
#
####### A) Create df
# Slopes df (employed in fig-learning B)
slope_list <- list(`Control-Cold` = slope_ControlCold,
                 `CORT-Cold` = slope_CORTCold,
                 `Control-Hot` = slope_ControlHot,
                 `CORT-Hot` = slope_CORTHot)
data_fig_learning_slopes <- do.call(rbind, lapply(names(slope_list), function(x) {
  data.frame(treatment = x, slopes = slope_list[[x]])
}))
# Intercepst df (only for fig-learning A)
intercept_list <- list(`Control-Cold` = int_ControlCold,
                 `CORT-Cold` = int_CORTCold,
                 `Control-Hot` = int_ControlHot,
                 `CORT-Hot` = int_CORTHot)
data_fig_learning_intercepts <- do.call(rbind, lapply(names(intercept_list), function(x) {
  data.frame(treatment = x, intercepts = intercept_list[[x]])
})) %>% 
  group_by(treatment) %>%
  summarize(int_sd = sd(intercepts),
            intercepts = mean(intercepts))
# df fig-learning A
data_fig_learning_slopesA <- data_fig_learning_slopes %>%
  group_by(treatment) %>%
  summarize(slope_sd = sd(slopes),
            slopes = mean(slopes),)
data_fig_learningA <- merge(data_fig_learning_slopesA, data_fig_learning_intercepts, by = "treatment")
treatment <- unique(data_fig_learningA$treatment)
fig_A_df <- data.frame()
for(t in treatment){ # Loop per treatment
  df <- data_fig_learningA %>%
    filter(treatment == t) %>%
    data.frame()
  # Variables selected
  m <- df$slopes
  u <- df$intercepts
  # Loop per treatment
  num_individuals <- length(u)
  for(x in 0:40){
    if (x == 0){
      sd <- df$int_sd
    } else {
      q25 <- df$slope_sd
    }
    value <- exp(u + m * x)
    temp_df <- data.frame(trial = rep(x, length(value)),
                          treatment = rep(t, length(value)),
                          errors = value,
                          sd = sd * value)
    fig_A_df <- rbind(fig_A_df, temp_df)
  }
}
# 
#### C) Plot the fig-learning A and fig-learning B separately 
fig_learningA <- plot_errorsday(fig_A_df)
fig_learningB <- plot_slopes(data_fig_learning_slopes)
#
#### D) Combine plots A and B to have fig-learning
fig_results_learning <- plot_grid(fig_learningA, fig_learningB, nrow = 1) +
  # Insert title for each plot
  annotate("text", x = 0.045, y = 0.935, label = "A", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.545, y = 0.935, label = "B", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.705, y = 0.89, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.68, y = 0.72, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.7, y = 0.55, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.69, y = 0.38, label = "*", hjust = 1, vjust = 1, size = 7)
#
ggsave("./output/figures/text/fig_results_errors.png", plot=fig_results_learning , width = 21, height = 7, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/text/fig_results_errors.png")
```

@@

Mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation were also not affected by prenatal conditions (see @fig-results_mit and table S2). We did not see any significant effects of sex and age on mitochondrial physiology either (see table S3 and tables S5-S9).

@@

### Declaration of AI use
We declare Chat GPT was used mainly for questions related to coding and data analyses. All other parts of the manuscript were written by the authors and those parts where AI was used were checked and modified accordingly by the authors.

@@

>>>>

Pablo Recio^1^,$\ddagger$ , Dalton C. Leibold^1^, Ondi L. Crino ^1,2^, Christopher R. Friesen^3,4^, Daniel W.A. Noble^1^

^1^ Division of Ecology and Evolution, Research School of Biology, The Australian National University, Canberra, ACT 2601, Australia  
^2^ Flinders University, College of Science and Engineering, Bedford Park, SA 5042, Australia  
^3^ University of Wollongong, Wollongong, NSW 2500, Australia  
^4^ Environmental Futures University of Wollongong, Wollongong, NSW 2500, Australia
$\ddagger$ Corresponding author: pablo.reciosantiago@anu.edu.au

@@

## Abstract
Early environmental factors like heat or stress hormones can impair learning through brain metabolic function, crucial for neural development and synaptic plasticity. However, whether early environments always result in cognitive impairment through changes in neural physiology is not well-established outside of a few model systems. Here, we investigated the effects of prenatal temperature and corticosterone (CORT) on brain mitochondrial activity and spatial learning in the delicate skink (_Lampropholis delicata_). We treated eggs with either CORT or a control vehicle and incubated at cold (23 ± 3°C) or hot (28 ± 3°C) temperatures. Juveniles were tested in a spatial learning task over 40 days after which mitochondrial function in the medial cortex was assessed. Despite among-individual variation in learning ability, mitochondrial physiology and spatial learning in _L. delicata_ remained robust to prenatal temperature and CORT exposure. No significant relationship was found between mitochondrial function and cognitive performance, contrary to predictions. Increased metabolic capacity correlated with higher ROS production but did not affect oxidative damage, possibly due to protective mechanisms. These findings highlight the physiological and cognitive resilience of _L. delicata_ to early-life challenges. Future research should explore whether this robustness extends to other brain regions, cognitive domains, and life stages.

@@

Learning - the acquisition and consolidation of new information - enables animals to create new associations between events, which can be essential for coping with environmental change [@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013]. However, the capacity to form new associations varies among individuals, potentially affecting their responses to environmental challenges [@ward2016ecological; @welklin2024spatial]. Learning faster may imply better exploitation of resources or responding more efficiently to novel threats [@ward2016ecological]. Consequently, individual differences in learning abilities can profoundly affect fitness, ultimately influencing population growth rates and stability [@ward2016ecological; @welklin2024spatial]. Therefore,understanding the mechanisms that drive these differences is crucial for predicting how populations will respond to novel conditions.

Factors like age, sex, or early-life conditions can have important effects on learning abilities [@szuran_water_1994; @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @carazo_sex_2014; @noble_age-dependent_2014]. Adverse developmental environments are particularly influential, as the brain is highly sensitive to environmental inputs during early life [@zhu_prenatal_2004]. For example, early experiences can alter neurotransmitter production [@amani2021perinatal], gene expression [@zhou2020effects], or brain structure [@amiel_effects_2017], with lasting effects on cognition. Among the physiological mechanisms underlying learning, mitochondrial activity is considered a key factor [@siegel1994basic; @du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. The neural processes involved in learning impose substantial energetic demands [@mcnay_decreases_2000; @mann_coupling_2021; @alexandrov_neuronal_2022], making efficient mitochondrial respiration essential [@du_dynamic_2009; @picard_mitochondria_2014; @picard_energetic_2018]. Additionally, learning depends on a dense network of functional neurons [@lefebvre_taxonomic_2011; @amiel_effects_2017], which can be compromised by excessive oxidative stress - a byproduct of mitochondrial metabolism [@finkel_oxidants_2000; @zhu_prenatal_2004; @du_dynamic_2009; @gong_chronic_2011; @hoffmann_mitochondrion_2018]. Studies on mammals have shown the pervasive effects of mitochondrial physiology on cognitive abilities [@zhu_prenatal_2004; @hara_presynaptic_2014; @cao2014ampk]. However, the extent to which these effects are generalisable to other taxa remains largely unexplored.

Since mitochondria are maternally inherited, maternal condition plays a fundamental role in shaping offspring mitochondrial activity [@picard_mitochondria_2014]. Maternal stress can also influence how mitochondria operate in offspring [@zhu_prenatal_2004]. Stressful situations experienced by mothers can elevate glucocorticoids (GCs) [@sapolsky_how_2000], which can affect developing embryos [@uller_sex-specific_2009] altering mitochondrial physiology through transgenerational effects [@picard_mitochondria_2014]. For instance, maternal stress has been shown to contribute significantly to oxidative stress in the brain of rats (_Rattus norvegicus_) with impacts on spatial learning abilities [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Temperature is also a significant source of maternal stress in ectotherms. Thermal environments outside the optimal range can elevate GCs in mothers, which can be passed to the offspring [see @Crino_2023]. Temperature can also directly influence offspring development, particularly during early life [@noble_developmental_2018; @crino2024eggs]. The prenatal thermal environment is crucial in shaping mitochondrial function, affecting energy metabolism and oxidative stress [@stier2022experimental; @crino2024eggs]. Thus, the combined effects of prenatal GCs and temperature may profoundly influence mitochondrial function, with important consequences for brain development and cognition. However, the extent to which prenatal GCs and temperature interact to shape cognitive abilities via mitochondrial physiology remains largely unknown outside of a few model species [@zhu_prenatal_2004; @haussmann_embryonic_2012; @cao2014ampk].

Here, we examined how prenatal temperature and corticosterone (CORT) — the primary GC in reptiles — affect brain mitochondrial physiology and spatial learning in the delicate skink (_Lampropholis delicata_). We hypothesized that prenatal CORT and temperature would influence learning abilities by impacting brain mitochondrial activity. Specifically, we predicted that prenatal CORT would decrease energy production while increasing oxidative damage [@costantini2011meta; @gong_chronic_2011; but see @crino2024eggs], especially if CORT can alter cellular components increasing ROS production without enhancing ATP synthesis or making them more vulnerable to oxidative damage. Similarly, we predicted that high temperatures would decrease mitochondrial efficiency [@zavorka_climate_2021; @crino2024eggs], but would reduce oxidative stress [@treidel2016temperature]. We further predicted that the combined effects of CORT and temperature would lead to complex interactions, with both factors negatively impacting mitochondrial efficiency but having opposite effects on oxidative stress. These effects would lead to differences in learning abilities, which could be affected by the balance between energy production and oxidative stress [@du_dynamic_2009; @picard_energetic_2018; @alexandrov_neuronal_2022]. By examining these interactions, we aim to clarify how prenatal environmental factors shape learning abilities through mitochondrial function, clarifying the mechanisms that mediate the role of early-life conditions on cognitive development.

@@

#### Experimental animals
Lizards came from a breeding colony established in the laboratory in 2019 from wild populations in Sydney, Australia. The colony consisted of 270 adults in groups of two males and four females. Eggs were collected from these groups between November 2023 to January 2024. After collecting the eggs, we treated them with CORT or vehicle control and incubated them under two different temperature regimes (see below). Clutch and egg identity were assigned immediately after egg collection, and the eggs were incubated in individual cups until hatching. Hatchlings were also kept in individual enclosures until the end of the experiment. For details on husbandry and breeding conditions see _Methods 1: Animal husbandry in Supplementary Material_.

@@

#### Spatial learning  
The spatial learning task involved training lizards to navigate a 6-arm maze to reach an exit connected to a transport box that allowed us to return the lizards to their enclosure without further contact. In each trial, lizards were placed by hand in the center of the maze and left to acclimatise for two minutes. During acclimatisation, the central area was surrounded by a yellow device mounted on a pulley system. At the start of each trial, this device was lifted to startle the lizard. If the lizard did not immediately choose an arm, it was gently prodded with a brush at the end of the tail. Once the lizard made a choice, the brush encouraged movement but without guiding it towards any specific arm. If the lizard did not choose correctly after 20 errors — at which point it typically ceased responding — the lizard was gently guided to the correct arm.

We assessed lizards' spatial learning using external cues for navigation while avoiding intra-maze cues. Subtle intra-maze cues were avoided by replacing the maze every three trials with one of four identical mazes. Each replacement preserved the correct arm’s orientation and the maze’s position within the room, ensuring consistency for each individual. Additionally, the maze was cleaned with 70% ethanol between trials to avoid the influence of chemical cues. The correct arm was randomly assigned to one of the six arms for each trial to control for potential side biases. We employed four maze orientations to counterbalance the number of lizards assigned to each orientation across treatments.

The task was repeated once daily for 40 consecutive days, and the number of errors made by each lizard was recorded. An error was defined as the lizard inserting its head into one of the incorrect arms with each incorrect choice, including consecutive choices of the same arm, counted as independent errors.

#### Brain mitochondrial function
Immediately after completing the tests, we quantified mitochondrial physiology in the brains of lizards. We euthanized lizards via intraperitoneal injection of 10 mg/kg of a 10 mg/mL alfaxan solution (a potent anesthetic), followed by decapitation. Before decapitation, we evaluated lizard righting responses and the pinching reflexes in one of the front limbs.

@@

The aliquots reserved for oxidative damage assays were stained with 10 µg/mL Hoechst 33342 Nuclear Viability Dye and 100 µM BODIPY 665/676 Lipid Peroxidation Sensor before cryopreservation. These dyes were used to measure cell viability and lipid peroxidation, respectively. The samples were then fixed in 1% Neutral-Buffered Formalin, washed, and preserved in a 1X Tris-EDTA solution with 10% DMSO at -20 °C. On the day of the oxidative damage assays, the samples were thawed, and DMSO was removed. Then, cell membranes were permeabilized in 200 µL of 1X PBS containing 20µM digitonin. Following permeabilization we stained the samples with 20µL of 70µM 8-OHdG Polyclonal Antibody - a marker of DNA damage - and allowed them to incubate overnight at 4 ºC. The following day we counterstained the cells with 20µL of 100 µg/mL H+G Goat Anti-Rabbit Conjugate Antibody with Alexa-Fluor 488 and analyzed the samples using the flow cytometer. Oxidative damage assays were performed within 6 months of the initial preparation of fresh samples.

Flow cytometry was performed on a 5-lasers flow cytometer (Becton Dickson LSRFortessa X-20) using the default wavelength filters and a high-throughput plate reader. Detectors and voltage settings for each assay were determined during pilot trials and remained consistent throughout the experiment. Data was processed using FlowJo (v. 10.1) software. We obtained the mean fluorescent intensity for mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation. For further details on the homogenization, staining, or flow cytometry assays, see _Methods 2: flow cytometry in Supplementary Material_. Sample sizes for DNA damage and lipid peroxidation were smaller than for the other variables because of a plater reader malfunction; however, data was missing randomly and we dealt with missing data in our analyses (see below). We validated that our homogenates contained neurons in a pilot study that used dyes specifically targeting neuronal nuclei (See _Brain validation in Supplementary Material_). This pilot study also ensured that our gating strategy identified these neurons using Flow cytometry.

#### Statistical analyses
We performed all analyses using the _brm_ package [@burkner2017brms], which fits Bayesian multilevel models with _Stan_ [@stan] with R version 4.4.0 [@R]. We ran a series of univariate and multivariate models to test the effects of early envinonment on each variable separately plus quantify the relationships between physiology and learning. All models consisted of four MCMC chains of 8000 iterations, with a warmup interval of 2000 iterations.

Univariate models were used to test the effects of early environment on each variable recorded: mitochondrial density, metabolic capacity, ROS, DNA damage, lipid peroxidation, and the number of errors as a measure of learning. In all the models, we included hormone treatment (CORT vs Control), incubation temperature (Cold vs Hot), and their interaction. Sex and age were included in preliminary models and excluded from the final models when they were not significant. For all univariate models, clutch identity was included as a random factor. Mitochondrial-related variables where log-transformed when necessary, and all were mean-centered and standardized by dividing them by two times the standard deviation [@gelman2008scaling]. These variables were considered to follow a normal distribution.

Learning was modeled as a function of trial, CORT, temperature, and their three-way interaction. In this model, we included the trial within each level of lizard identity as a random slope. The error structure was modelled using a negative binomial distribution with a logit link function [negbinomial(link = "log")]. Otherwise, the procedure was as in the other models.

We used the posterior distributions of parameters from these models to test for differences between treatments. Learning slopes were obtained using the ‘trial’ estimates and its interaction with hormone and temperature treatments. Slope estimates less than 0 provide evidence of learning. _pMCMC_ was used to test the hypothesis that posterior distributions of slopes and slope contrasts differed from zero. We considered an effect statistically significant when _pMCMC_ <0.05.

We fitted a multivariate structural equation model (SEM) using _brms_ to explore direct and indirect links between mitochondrial function and learning. The model was structured based on theoretical expectations shown in @fig-sem. Any missing data was imputed during model fitting using data augmentation [@noble2021planned], but this was largely restricted to DNA damage and lipid peroxidation. To obtain a measure of learning performance we extracted the posterior distribution of individual learning slopes (i.e., changes in error over time) by including a random trial slope for each lizard. Learning slopes and the rest of variables, were standardized as before (see above). We assumed a Gaussian error distribution for all the variables in this model. Factors found to be non-significant in the univariate models were excluded. Clutch identity remained a random factor in our SEM model. Direct, indirect, and total effects were derived from posterior estimates [@kline2005principles] (see table S4).

@@

We started with 80 lizards - 20 per treatment - from `r total_clutches` clutches. However, due to natural mortality, only 79 lizards were included in the learning analyses, and 78 in the mitochondrial analyses. We provide final sample sizes in all our figures.

@@

```{r, fig-learning}
#| label: fig-learning
#| fig.cap: "Model estimates for learning analyses. (a) the predicted number of errors over trials. The lines represent the mean predicted number of errors for each trial, and the shaded areas indicate the standard error of the mean; both were obtained using the slope and intercept estimates from the posterior distributions. (b) shows the distribution of the estimates of slopes per each treatment. The x-axis represents the slope estimate, and in the y-axis are the density of the estimates. Points and bars represent the mean and standard error of the estimated slopes, respectively. Dashed lines indicate value 0. The different colors in both panels indicate the different treatments. Asterisks in (b) indicate significant differences from 0."
source(here("R", "func.R"))
#
#
####### A) Create df
# Slopes df (employed in fig-learning B)
slope_list <- list(`Control-Cold` = slope_ControlCold,
                 `CORT-Cold` = slope_CORTCold,
                 `Control-Hot` = slope_ControlHot,
                 `CORT-Hot` = slope_CORTHot)
data_fig_learning_slopes <- do.call(rbind, lapply(names(slope_list), function(x) {
  data.frame(treatment = x, slopes = slope_list[[x]])
}))
# Intercepst df (only for fig-learning A)
intercept_list <- list(`Control-Cold` = int_ControlCold,
                 `CORT-Cold` = int_CORTCold,
                 `Control-Hot` = int_ControlHot,
                 `CORT-Hot` = int_CORTHot)
data_fig_learning_intercepts <- do.call(rbind, lapply(names(intercept_list), function(x) {
  data.frame(treatment = x, intercepts = intercept_list[[x]])
})) %>% 
  group_by(treatment) %>%
  summarize(int_sd = sd(intercepts),
            intercepts = mean(intercepts))
# df fig-learning A
data_fig_learning_slopesA <- data_fig_learning_slopes %>%
  group_by(treatment) %>%
  summarize(slope_sd = sd(slopes),
            slopes = mean(slopes),)
data_fig_learningA <- merge(data_fig_learning_slopesA, data_fig_learning_intercepts, by = "treatment")
treatment <- unique(data_fig_learningA$treatment)
fig_A_df <- data.frame()
for(t in treatment){ # Loop per treatment
  df <- data_fig_learningA %>%
    filter(treatment == t) %>%
    data.frame()
  # Variables selected
  m <- df$slopes
  u <- df$intercepts
  # Loop per treatment
  num_individuals <- length(u)
  for(x in 0:40){
    if (x == 0){
      sd <- df$int_sd
    } else {
      q25 <- df$slope_sd
    }
    value <- exp(u + m * x)
    temp_df <- data.frame(trial = rep(x, length(value)),
                          treatment = rep(t, length(value)),
                          errors = value,
                          sd = sd * value)
    fig_A_df <- rbind(fig_A_df, temp_df)
  }
}
# 
#### C) Plot the fig-learning A and fig-learning B separately 
fig_learningA <- plot_errorsday(fig_A_df)
fig_learningB <- plot_slopes(data_fig_learning_slopes)
#
#### D) Combine plots A and B to have fig-learning
fig_results_learning <- plot_grid(fig_learningA, fig_learningB, nrow = 1) +
  # Insert title for each plot
  annotate("text", x = 0.045, y = 0.935, label = "A", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.545, y = 0.935, label = "B", hjust = 1, vjust = 1, size = 7, fontface = "bold") +
  annotate("text", x = 0.705, y = 0.89, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.68, y = 0.72, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.7, y = 0.55, label = "*", hjust = 1, vjust = 1, size = 7) +
  annotate("text", x = 0.69, y = 0.38, label = "*", hjust = 1, vjust = 1, size = 7)
#
ggsave("./output/figures/text/fig_results_errors.png", plot=fig_results_learning , width = 21, height = 7, units = "cm", dpi = 600)
knitr::include_graphics("./output/figures/text/fig_results_errors.png")
```

@@

Mitochondrial density, metabolic capacity, ROS, DNA damage, and lipid peroxidation were also not affected by prenatal conditions (see @fig-results_mit and table S2). We did not find effects of sex and age on mitochondrial physiology (see table S3 and tables S5-S9).

@@

### Declaration of AI use
We declare Chat GPT-4 was used for questions related to code checking when there were unknown errors, but we did not rely on AI for the creation of any of the content in the manuscript, which was written fully by the authors. Last time Chat GPT-4 was used was February 2025.

@@

#### Brain validation
